Local Search Approximation Algorithms for Partitional and Hierarchical Clustering

### 2015-05-10 tex/project.tex

An objective function that is somewhat in between $k$-means and $k$-center is to minimize the sum of distances from each point to the median point of its cluster, called $k$-median or $k$-medoids clustering.  There are two common heuristic algorithms for this problem. The first is a local search algorithm called partition around medoids (PAM), which starts with $k$ randomly chosen data points as centers, and looks to swap one of the centers with another data point to improve the objective value.  This repeats until no more improving swaps exist \cite{ESL}. 
Another common heuristic (which we shall call $k$-medoids throughout this paper), which is more computationally efficient than PAM \cite{Park}, is a varitation on Lloyd's $k$-means algorithm.  Starting with randomly chosen centers, we assign each data point to its closest center, and then update the center within each cluster \cite{Park}. The algorithm repeats this until the solution becomes stable.  Additionally, Velmurugan and Santhanam showed that on uniform and normally distributed data, $k$-medoids was empirically faster than $k$-means \cite{Velmurugan}.

@@

The data sets tested against were:
\begin{description}
\item[Gaussian] Synthetic $k$-center Gaussian data, generated in Python. In particular, Gauss\textunderscore d \textunderscore k \textunderscore e was generated by first choosing k random centers in $[-1,1]^d$. Around each center, we then generated $(1000/k) \mathrm{Unif}[0,1]$ Gaussian points of which $e$ percent had standard deviation 0.25 (noisy) and $1-e$ percent had standard deviation $0.05$. 
\item[OR $p$-median Library] From a collection of test data sets created by J. E. Beasley with data sets of size $100, 200, \ldots, 900$ and varying values of $k$ \cite{Beasley}. 
\item[UCI Iris] \cite{Iris} Contains 4 attributes on 150 instances of flowers, each of with belongs to one of three species of iris.  One of the species is linearly separable from the others, but the other two are not.  One of the most used data sets in pattern recognition.
\item[UCI Soybean (Small)] \cite{Soybean} 47 instances, with 35 attributes data set of soybean disease from UCI.
\end{description}

\begin{figure}[h]
\begin{tabular}{ | l | l | l | l | l | }
\hline
	Data Set Name &Optimal Objective & $k$ & $n$ & dim   \\ \hline
	Soybean & 113.8 & 4 & 47 & 35 \\ 
	Iris & 98.2 & 3 & 150 & 4 \\ 
	Gauss\_5\_5\_5 & 195.4 & 5 & 388 & 5 \\ 
	Gauss\_5\_10\_5 & 270 & 10 & 544 & 5 \\ 
	Gauss\_5\_50\_5 & 178.1 & 50 & 408 & 5 \\ 
	pmed1 & 5819 & 5 & 100 & \  \\ 
	\hline
\end{tabular}
\caption{Some specifics about our datasets}
\end{figure}

Tests were conducted using Python 2.7.9 on a 64-bit Windows platform, on a 4 core Intel processor.

\begin{figure*}[h]
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Time & Time & Time & Time & Time & Time & Time & Time & Time \\ \hline
	&K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap  \\ \hline
	Soybean & 0.015 & 0.061 & 0.002 & 1.179 & 0.026 & 0.016 & 0.048 & 0.049 & 0.081 \\ 
	Iris & 0.013 & 0.023 & 0.003 & 9.788 & 0.125 & 0.102 & 0.212 & 0.229 & 0.414 \\ 
	Gauss\_5\_5\_5 & 0.025 & 0.056 & 0.032 & 75.262 & 1.159 & 1.506 & 2.188 & 2.382 & 3.71 \\ 
	Gauss\_5\_10\_5 & 0.054 & 0.095 & 0.109 & 134.137 & 6.834 & 5.407 & 9.243 & 9.428 & 15.96 \\ 
	Gauss\_5\_50\_5 & 0.137 & 0.196 & 0.302 & 97.336 & 27.856 & 21.446 & 32.407 & 39.614 & 44.69 \\ 
	pmed1 & N/A & N/A & 0.003 & 5.042 & 0.084 & 0.09 & 0.155 & 0.161 & 0.255 \\ \hline
\end{tabular}

\caption{Running Times For Algorithms}
\end{figure*}

\begin{figure*}[h]
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI  \\ \hline
	& K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap   \\ \hline
	Soybean & 0.69 & 0.18 & 0.55 & 0.61 & 0.71 & 0.73 & 0.72 & 0.71 & 0.71  \\ 
	Iris & 0.75 & 0.76 & 0.69 & 0.7 & 0.77 & 0.76 & 0.76 & 0.76 & 0.76  \\ 
	Gauss\_5\_5\_5 & 0.74 & 0.71 & 0.69 & 0.66 & 0.74 & 0.75 & 0.74 & 0.74 & 0.74  \\ 
	Gauss\_5\_10\_5 & 0.78 & 0.74 & 0.68 & 0.58 & 0.73 & 0.72 & 0.74 & 0.74 & 0.73  \\ 
	Gauss\_5\_50\_5 & 0.65 & 0.6 & 0.62 & 0.57 & 0.62 & 0.62 & 0.63 & 0.63 & 0.62  \\ \hline
\end{tabular}

\caption{AMI For Algorithms}
\end{figure*}

\begin{figure*}[h]
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Approx & Approx & Approx & Approx & Approx & Approx & Approx  \\ \hline
	& k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps+ & 3-Swap   \\ \hline
	Soybean & 1.17 & 1.2 & 1.03 & 1.01 & 1.02 & 1.01 & 1.02  \\ 
	Iris & 1.08 & 1.37 & 1.02 & 1 & 1.01 & 1 & 1.01  \\ 
	Gauss\_5\_5\_5 & 1.07 & 1.44 & 1.03 & 1 & 1.03 & 1.01 & 1.03  \\ 
	Gauss\_5\_10\_5 & 1.08 & 1.39 & 1.02 & 1.01 & 1.02 & 1.01 & 1.02  \\ 
	Gauss\_5\_50\_5 & 1.09 & 1.27 & 1.03 & 1.02 & 1.03 & 1.01 & 1.03  \\ 
	pmed1 & 1.11 & 1.41 & 1.03 & 1.01 & 1.03 & 1.01 & 1.02  \\ \hline
\end{tabular}

\caption{Approximation Coeffiecients For Algorithms}
\end{figure*}

In terms of computation time, our proposed algorithms were much slower than $k$-means, the spectral algorithm, or $k$-medoids. In general, the multiswaps algorithm was around two orders of magnitude slower with a small multiplicative increase as $p$ increased, and the supermodular algorithm was around four orders of magnitude slower, likely due to the extra sampling the supermodular algorithm must perform.

@@

When increasing $p$ for the multiswaps algorithm, we did observe slight increases in computation time, but the times remained on the same order of magnitude. The increase in $p$ did not seem to have any significant effect on the AMI measure nor approximation ratio. This may be due to the fact the algorithm for $p=1$ performed quite well already with an approximation ratio very close to 1 that beat out $k$-medoids. In addition, PAM had the highest AMI measures for the iris and soybean data sets. The AMI measures for the Gaussian data set were higher than those of $k$-medoids in general but equivalent or only slightly weaker than those of the spectral method or $k$-means. Adding in a few rounds of $k$-medoids at the end of our multiswaps algorithm did not seem to have much of an effect on the AMI measure but did boost the approximation ratio, sometimes finding an optimal solution.

It is interesting to note that higher dimensional Gaussian data seemed easier for the algorithms to classify along with smaller values of $k$ and lower percentage of noisy data, which makes sense based on our intuition of clustering. For the OR library instances, we saw that the approximation ratio tended to increase as $k$ grew larger but not as $n$ grew.

Overall, the results of our experiments do not point to the fact that these proposed approximation algorithms can add any advantage over current methods. However, the results do point to the fact that we can use sampling to run PAM with lower computational cost while still maintaining good performance accuracy both in terms of AMI measure and approximation ratios. Adding in a few rounds of $k$-medoids seems also to be a good method to refine a given solution towards a local optimum.

\section{Hierarchical Clustering Approximation Algorithms}
Suppose we have a set $X$ and a distance metric $d(i,j)$ for all $i,j \in X$. In the hierarchical setting, we will represent a solution to the $k$-median problem as a pair $(S, a)$ where $S \subseteq X$ such that $|S| \leq k$ and $a$ is an assignment function that maps each $i \in X$ to a center in $S$. Therefore, for any $i \in S$, $a^{-1}(i)$ will be all points in the cluster centered at $i$. We define
\[ \mathrm{cost}(S,a) := \sum_{i \in X} d(i, a(i)) . \]

Given solution pairs $(S_1, a_1)$ and $(S_2, a_2)$, we say that $(S_1, a_1)$ is \emph{nested} in $(S_2, a_2)$ if 
\begin{enumerate}
\item $S_1 \subset S_2$, 
\item  $\forall  j \in X$, if $a_2(j) \in S_1$, then $a_1(j) = a_2(j)$, and 
\item $\forall j, k \in X$, if $a_2(j) = a_2(k)$, then $a_1(j) = a_1(k)$.
\end{enumerate}
 A solution to the hierarchical clustering problem is therefore a chain of nested solutions. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the algorithm.  
Consider any $k$-median algorithm $\mathcal{A}$, such as the algorithms above. After running $\mathcal{A}$ for all potential values of $k$, Lin et al.~\cite{Lin} introduce  augmentation step that will nest these clusterings so that they fit into the hierarchical framework and only increase the costs by a bounded factor.

First, we set $i=1$, $S_0 = \emptyset$, $\beta = 3+\sqrt{3}$, and $\beta_0 = 1$ (for the randomized algorithm we switch $\beta = 6.355$ and set $\beta_0 = \beta^X$, where $X$ is chosen from a uniform distribution on $[0,1]$). Let $V_i$ be the solutions generated from running $\mathcal{A}$ for $k=n-i+1$ for $i=1, \ldots, n$. The cost of these solutions is determined by assigning each point to the nearest cluster center and is scaled so that the minimum cost for $k<n$ is at least one. We then order these solutions according to their cost into buckets of the form $[0,0], (\beta_0, \beta_0 \beta], (\beta_0 \beta, \beta_0 \beta^2], \ldots $.

From each non-empty bucket, pick the solution with the smallest number of cluster centers and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r = V_n$. We then inductively build up our nested solutions in the following manner. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we define $S_{i+1}$ by iteratively adding the closest point $j \in S_{i-1}$ to $i$ for all $i \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

@@

We continue this nesting until $|S_i| = 1$ and we have reached the top level of the hierarchy. After forming this hierarchical clustering, for any $k \in 1, \ldots, n$ \cite{Lin} guarantee that the cost of $S_i$ is at most $20.71$ times the cost of $V_k$ where $|S_{i-1}| \geq n-k+1 \geq |S_{i}|$ (in the randomized case the expected cost of $S_i$ is at most $10.03$ times the cost of $V_k$).

We test this hierarchical framework for both the multiswap and supermodular local search algorithms above.

\subsection{Implementation and Results}
We implemented the nesting hierarchical framework above in Python offering options for both the randomized and deterministic versions. As a black-box $k$-median solver, we used $k$-medoids and the 1-swap+ algorithms above. We compared the results to hierarchical agglomerative clustering with both average linkage and Ward's linkage (implemented in the scipy package in Python).

We also tested a more computationally friendly heuristic of the hierarchical framework which just buckets solutions for values of $k$ that are a power of two. This yields many fewer $k$-median instances to solve and reduced the run-time significantly. This was motivated by the fact that with the difference between various $k$-median solutions will be larger when $k$ is small so we expect our sampling to still represent much of the underlying structure.

For each data set and each clustering method, we report the approximation ratio, computation time, and FScore measure.  FScore is a measure that looks at the clusters represented on each level and how they represent the true clustering in terms of precision and recall. See \cite{Larsen} for more information.

In general, the hierarchical framework performed significantly better than both methods of agglomerative clustering in terms of FScore measure even for our logarithmic version. The FScore measures were very low for the agglomerative clustering methods, often around 0.25, but shot up to closer to the AMI measures we saw in flat clustering for our implemented algorithms.

For $k$-medoids in the non-logarithmic version and PAM in the logarithmic version, the hierarchical framework increased the computation time by three orders of magnitude, whereas this was decreased to two orders difference for $k$-medoids logarithmic version. Interestingly, the approximation factors did not seem to change between the randomized and deterministic versions for $k$-medoids. The approximation ratios seemed best as $n$ increased or as the dimension increased. In particular, for the Gaussian data in 7 dimensions, the approximation ratio was often around 3. Recall that this approximation measures the constructed $k$-median solution on every level against an optimal one so we expect the ratio to not come close to 1 in many cases.

\section{Conclusion and Future Work}
For partitional clustering, it seems that the approximation algorithms in the literature are in general slower and cannot improve much on the current heuristics used despite their approximation guarantees. This was not true in the hierarchical clustering case where we saw significant improvement in the FScore measures for our implemented methods and overall good approximation ratios where none were had before. In particular, our logarithmic nesting algorithm seems to provide the best balance between performance and computation.

In the future, we hope to explore how linear programming based methods perform in this context as well and to implement our own agglomerative clustering so we can test it on the OR library instances. Lastly, when nesting solutions in the logarithmic framework, we could use agglomerative clustering to make smarter local merges.

\section{Acknowledgements}
We would like to acknowledge David P. Williamson and Karthik Sridharan for their valuable feedback on this project.

>>>>

An objective function that is somewhat in between $k$-means and $k$-center is to minimize the sum of distances from each point to the median point of its cluster, called $k$-median or $k$-medoids clustering.  There are two common heuristic algorithms for this problem. The first is a local search algorithm called partition around medoids (PAM), which starts with $k$ randomly chosen data points as centers, and looks to swap one of the centers with another data point to improve the objective value.  This repeats until no more improving swaps exist \cite{ESL}. 
Another common heuristic (which we shall call $k$-medoids throughout this paper), which is more computationally efficient than PAM \cite{Park}, is a varitation on Lloyd's $k$-means algorithm.  Starting with randomly chosen centers, we assign each data point to its closest center, and then update the center within each cluster \cite{Park}. The algorithm repeats this until the solution becomes stable.

@@

The data sets tested against were:
\begin{description}
\item[Gaussian] 27 synthetic $k$-center Gaussian data sets, generated in Python. In particular, Gauss\textunderscore d\textunderscore k\textunderscore e was generated by first choosing k random centers in $[-1,1]^d$. Around each center, we then generated $(1000/k) \cdot \mathrm{Unif}[0,1]$ Gaussian points of which $e$ percent had standard deviation 0.25 (noisy) and $1-e$ percent had standard deviation $0.05$. 
\item[OR $p$-median Library] From a collection of test data sets created by \cite{Beasley} with 40 test sets of size $100, 200, \ldots, 900$ and varying values of $k$. 
\item[UCI Iris] \cite{Iris} Contains 4 attributes on 150 instances of flowers, each of with belongs to one of three species of iris.  One of the species is linearly separable from the others, but the other two are not.  One of the most used data sets in pattern recognition.
\item[UCI Soybean (Small)] \cite{Soybean} 47 instances, with 35 attributes data set of soybean disease from UCI.
\end{description}

\begin{figure}[h]
\begin{tabular}{ | l | l | l | l | l | }
\hline
 Dataset Name & Opt Obj & $k$ & $n$ & dim \\ \hline
Soybean & 114 & 4 & 47 & 35 \\ 
Iris & 98 & 3 & 150 & 4 \\ 
Gauss\_5\_5\_5 & 195.4 & 5 & 388 & 5 \\ 
Gauss\_5\_10\_5 & 270 & 10 & 544 & 5 \\ 
Gaussian avg & 241 & 22 & 507 & 5 \\ 
pmed5 & 1355 & 33 & 100 & \  \\ 
pmed20 & 1789 & 133 & 400 & \  \\ 
% pmed40 & 5128 & 90 & 900 & \  \\ 
pmedian avg & 5535 & 46 & 460 & \  \\ \hline
 \end{tabular}
\caption{Some specifics about our datasets}
\end{figure}

Tests were conducted using Python 2.7.9 on a 64-bit Windows platform, on a 4 core Intel processor. Results are given in Figure 2  below with some results averaged due to space constraints.

\begin{figure*}[h]
\label{fig:flat_results}
\small
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Time & Time & Time & Time & Time & Time & Time & Time & Time \\ \hline
	&K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap  \\ \hline
	Soybean & 0.015 & 0.061 & 0.002 & 1.179 & 0.026 & 0.016 & 0.048 & 0.049 & 0.081 \\ 
	Iris & 0.013 & 0.023 & 0.003 & 9.788 & 0.125 & 0.102 & 0.212 & 0.229 & 0.414 \\ 
	Gauss\_5\_5\_5 & 0.025 & 0.056 & 0.032 & 75.262 & 1.159 & 1.506 & 2.188 & 2.382 & 3.71 \\ 
	Gauss\_5\_10\_5 & 0.054 & 0.095 & 0.109 & 134.137 & 6.834 & 5.407 & 9.243 & 9.428 & 15.96 \\ 
	%Gauss\_5\_50\_5 & 0.137 & 0.196 & 0.302 & 97.336 & 27.856 & 21.446 & 32.407 & 39.614 & 44.69 \\ 
	Gaussian avg & 0.073 & 0.128 & 0.182 & 144.455 & 13.289 & 13.021 & 18.749 & 18.696 & 24.592 \\ 
	pmed5 & 0.038 & 8.298 & 0.481 & 0.622 & 0.653 & 0.748 & 0.939 & \  & \  \\ 
	pmed20 & 1.795 & 174.951 & 52.208 & 44.979 & 68.554 & 68.231 & 79.288 & \  & \  \\ 
	% pmed40 & 3.527 & 849.852 & 251.259 & 264.867 & 250.412 & 391.738 & 350.235 & \  & \  \\ 
	pmedian avg & 0.807 & 203.308 & 40.386 & 40.347 & 48.264 & 56.76 & 56.19 & \  & \  \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI  \\ \hline
	& K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap   \\ \hline
	Soybean & 0.686 & 0.178 & 0.55 & 0.612 & 0.708 & 0.732 & 0.717 & 0.708 & 0.712 \\ 
	Iris & 0.748 & 0.757 & 0.69 & 0.7 & 0.772 & 0.760 & 0.763 & 0.757 & 0.761 \\ 
	Gauss\_5\_5\_5 & 0.741 & 0.713 & 0.687 & 0.657 & 0.742 & 0.749 & 0.741 & 0.744 & 0.737  \\ 
	Gauss\_5\_10\_5 & 0.783 & 0.736 & 0.677 & 0.575 & 0.729 & 0.725 & 0.737 & 0.737& 0.729  \\ 
	%Gauss\_5\_50\_5 & 0.646 & 0.596 & 0.621 & 0.570 & 0.622 & 0.619 & 0.627 & 0.631 & 0.625  \\ 
	Gaussian avg & 0.777 & 0.744 & 9.112 & 0.623 & 0.76 & 0.761 & 0.762 & 0.761 & 0.76 \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Approx & Approx & Approx & Approx & Approx & Approx & Approx  \\ \hline
	& k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps+ & 3-Swap   \\ \hline
	Soybean & 1.167 & 1.199 & 1.027 & 1.012 & 1.022 & 1.012 & 1.024 \\ 
	Iris & 1.081 & 1.375 & 1.016 & 1.004 & 1.013 & 1.004 & 1.009 \\ 
	Gauss\_5\_5\_5 & 1.070 & 1.436 & 1.030 & 1.002 & 1.027 & 1.005 & 1.026  \\ 
	Gauss\_5\_10\_5 & 1.080 & 1.389 & 1.017 & 1.011 & 1.018 & 1.009 & 1.017  \\ 
	%Gauss\_5\_50\_5 & 1.0887 & 1.272 & 1.030 & 1.020 & 1.027 & 1.014 & 1.033  \\ 
	Gaussian avg & 1.064 & 1.416 & 1.025 & 1.009 & 1.023 & 1.008 & 1.024 \\ 
	pmed5 & 1.373 & 1.681 & 1.06 & 1.026 & 1.071 & 1.033 & 1.051 \\ 
	pmed20 & 1.501 & 1.73 & 1.079 & 1.032 & 1.076 & 1.027 & 1.073 \\ 
	%pmed40 & 1.187 & 1.423 & 1.046 & 1.017 & 1.043 & 1.019 & 1.039 \\ 
	pmedian avg & 1.203 & 1.456 & 1.041 & 1.017 & 1.039 & 1.016 & 1.037 \\ \hline
\end{tabular}

\caption{Average Computation Time (sec), AMI measure, and Approximation Coeffiecients for Partitional Clustering Algorithms}
\end{figure*}

In terms of computation time, our proposed algorithms were much slower than $k$-means, the spectral algorithm, or $k$-medoids. In general, the multiswaps algorithm was around two orders of magnitude slower with a small multiplicative increase as $p$ increased, and the supermodular algorithm was around four orders of magnitude slower, likely due to the extra potential function sampling the supermodular algorithm must perform.

@@

When increasing $p$ for the multiswaps algorithm, we did observe slight increases in computation time, but the times remained on the same order of magnitude. The increase in $p$ did not seem to have any significant effect on the AMI measure nor approximation ratio. This may be due to the fact the algorithm for $p=1$ performed quite well already with an approximation ratio very close to 1. In addition, PAM had the highest AMI measures for the iris and soybean data sets. The AMI measures for the Gaussian data set were higher than those of $k$-medoids in general but equivalent or only slightly weaker than those of the spectral method or $k$-means. The heuristic method of adding in a few rounds of $k$-medoids at the end of our multiswaps algorithm did not have much effect on the AMI measure but did boost the approximation ratio, sometimes finding an optimal solution.

It is interesting to note that higher dimensional Gaussian data, with smaller values of $k$ and a lower percentage of noisy data, was easier for the algorithms to classify, which makes sense based on our intuition of clustering. For the OR library instances, we saw that the approximation ratio tended to increase as $k$ grew larger but not as $n$ grew.

Overall, the results of our experiments do not suggest that the proposed approximation algorithms provide any advantage over current well known methods. However, the results do suggest that we can use sampling to run PAM with lower computational cost while still maintaining good performance accuracy in terms of AMI and approximation ratio.

\section{Hierarchical Clustering Approximation Algorithms}
Turning now to the hierarchical clustering setting, given a set $X$ and a distance metric $d(i,j)$ for all $i,j \in X$, we will represent a solution to the $k$-median problem as a pair $(S, a)$ where $S \subseteq X$ such that $|S| \leq k$ and $a$ is an assignment function that maps each $i \in X$ to a center in $S$. Therefore, for any $i \in S$, $a^{-1}(i)$ will be all points in the cluster centered at $i$. We define $\mathrm{cost}(S,a) := \sum_{i \in X} d(i, a(i))$.

Given solution pairs $(S_1, a_1)$ and $(S_2, a_2)$, we say that $(S_1, a_1)$ is \emph{nested} in $(S_2, a_2)$ if the following conditions hold.
\begin{enumerate}
	\item $S_1 \subset S_2$, 
	\item $\forall  j \in X$, if $a_2(j) \in S_1$, then $a_1(j) = a_2(j)$, 
	\item $\forall j, k \in X$, if $a_2(j) = a_2(k)$, then $a_1(j) = a_1(k)$.
\end{enumerate}
 A solution to the hierarchical clustering problem is therefore a chain of nested solutions. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the hierarchy.  
Consider any $k$-median algorithm $\mathcal{A}$. After running $\mathcal{A}$ for all potential values of $k$, Lin et al.~\cite{Lin} introduce an augmentation step that will nest these clusterings into the hierarchical framework and only increase costs by a bounded factor.

First, set $i=1$, $S_0 = \emptyset$, $\beta = 3+\sqrt{3}$, and $\beta_0 = 1$ (for the randomized algorithm we switch $\beta = 6.355$ and set $\beta_0 = \beta^X$, where $X$ is chosen from a uniform distribution on $[0,1]$). Let $V_i$ be the solutions generated from running $\mathcal{A}$ for $k=n-i+1$ for $i=1, \ldots, n$. The cost of these solutions is determined by assigning each point to the nearest cluster center and is scaled so that the minimum cost for $k<n$ is at least one. We then order these solutions according to their cost into buckets of the form $[0,0], (\beta_0, \beta_0 \beta], (\beta_0 \beta, \beta_0 \beta^2], \ldots $.

From each non-empty bucket, pick the solution with the smallest number of cluster centers and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r = V_n$. We then inductively build our nested solutions. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we define $S_i$ by iteratively adding the closest point $j \in S_{i-1}$ to $i$ for all $i \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

@@

We continue this nesting until $|S_i| = 1$, reaching the top level of the hierarchy. After forming this hierarchical clustering, for any $k \in 1, \ldots, n$ \cite{Lin} guarantee that the cost of $S_i$ is at most $20.71$ times the cost of $V_k$ where $|S_{i-1}| \geq n-k+1 \geq |S_{i}|$ (in the randomized case the expected cost of $S_i$ is at most $10.03$ times the cost of $V_k$).

\subsection{Implementation and Results}
We implemented the nesting hierarchical framework above in Python, offering options for both the randomized and deterministic versions. As a black-box $k$-median solver, we used $k$-medoids and the 1-swap+ algorithms, discussed in the partitional clustering section above. We compared the results to hierarchical agglomerative clustering with both average linkage and Ward's linkage (implemented in the scipy package in Python).

We also tested a more computationally friendly heuristic of the hierarchical framework which only buckets solutions for values of $k$ that are a power of two. This yields many fewer $k$-median instances to solve, reducing the run-time significantly. This was motivated by the fact that the difference between various $k$-median solutions will be larger when $k$ is small so we expect our sampling to still represent much of the underlying structure.

For each data set discussed previously, and each clustering method, we report the approximation ratio, computation time, and FScore measure.  FScore is a measure that looks at the clusters represented on each level and how they represent the true clustering in terms of precision and recall. See \cite{Larsen} for more information.  Results are in Figure 3.

\begin{figure*}[h]
\label{fig:hier_results}
\small
\begin{tabular}{ | l | l | l | l | l | l | l | }
\hline
	   & Time & Time & Time & Time & Time & Time \\ \hline
	Dataset & Ave Linking & Ward Linking & K-med & K-med logn ran & k-med ran & PAM logn ran \\ \hline
	Soybean & 0.0002 & 0.0002 & 0.2 & 0 & 0.2 & 0.2 \\ 
	Iris & 0.0023 & 0.003 & 3.2 & 0.2 & 3.8 & 5.1 \\ 
	Gauss\_5\_5\_5 &  & & & & & \\
	Gauss\_5\_10\_5 & & & & & & \\
	%Gauss\_5\_50\_5 &  & & & & & \\
	Gaussian Average & 0.0858 & 0.0853 & 354.5 & 7.5 & 369.9 & 423.3 \\ 
	pmed5 & 1.2 & 0.1 & 1.3 & 1.7 & \  & \  \\ 
	pmed25 & 279.3 & 5 & 306.7 & 312.5 & \  & \  \\ 
	pmedian avg & 87.4 & 2 & 94.5 & 101.6 & \  & \  \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | }
\hline
	   & F-Score & F-Score & F-Score & F-Score & F-Score & F-Score \\ \hline
	Dataset & Ave Linking & Ward Linking & K-med & K-med logn ran & k-med ran & PAM logn ran \\ \hline
	Soybean & 0.6 & 0.5 & 0.8 & 0.6 & 0.7 & 0.6 \\ 
	Iris & 0.5 & 0.5 & 0.9 & 0.8 & 0.8 & 0.7 \\ 
	Gauss\_5\_5\_5 &  & & & & & \\
	Gauss\_5\_10\_5 &  & & & & & \\
	%Gauss\_5\_50\_5 &  & & & & & \\
	Gaussian Average & 0.4 & 0.4 & 0.6 & 0.6 & 0.6 & 0.6 \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | }
\hline
	   & Approx & Approx & Approx & Approx \\ \hline
	Dataset & K-med & K-med logn ran & k-med ran & PAM logn ran \\ \hline
	Soybean & 1.6 & 2.3 & 1.7 & 2 \\ 
	Iris & 2560631.5 & 2550279.6 & 2787312 & 1623132.1 \\ 
	Gauss\_5\_5\_5 &  & & &  \\
	Gauss\_5\_10\_5 &  & & &  \\
	%Gauss\_5\_50\_5 &  & & &  \\
	Gaussian Average & 5.2 & 7.3 & 5.5 & 4.7 \\ 
	pmed5 & 11.4 & 51.1 & 16.3 & 12 \\ 
	pmed25 & 6.6 & 13.4 & 5.4 & 13.2 \\ 
	pmedian avg & 10.3 & 19.8 & 10.5 & 8.6 \\ \hline
\end{tabular}

\caption{Average Computation Time (sec), FScore measure, and Approximation Coeffiecients for Hierarchical Clustering Algorithms}
\end{figure*}

In general, the hierarchical framework performed significantly better than both methods of agglomerative clustering in terms of FScore measure, even for our logarithmic version. The FScore measures were very low for the agglomerative clustering methods, often around 0.25, but were much closer to the AMI measures we saw in flat clustering for our implemented algorithms.

For $k$-medoids in the non-logarithmic version and PAM in the logarithmic version, the hierarchical framework increased the computation time by three orders of magnitude, and increased by two orders of magnitude for $k$-medoids in the logarithmic version. Interestingly, the approximation factors did not change significantly between the randomized and deterministic versions for $k$-medoids. The approximation ratios were strongest as $n$ increased or as the dimension increased. In particular, for the Gaussian data in 7 dimensions, the approximation ratio was often around 3.  The high approximation ratio for the iris data set was due to the fact that the optimal solution for $k = n-1, n-2$ was very close to zero, and our heuristic $k$-median algorithm only sometimes picked up on these solutions. The approximation factor at lower levels was more consistent with the other results.

\section{Conclusion and Future Work}
For partitional clustering, we found that the approximation algorithms in the literature are in general slower and cannot improve much on current heuristics used despite their approximation guarantees. This was not true in the hierarchical clustering case where we saw significant improvement in FScore measures for our implemented methods and overall good approximation ratios where none were had before. In particular, our logarithmic nesting algorithm seems to provide the best balance between performance and computational efficiency.

In the future, we hope to explore how linear programming based methods perform in both the flat and hierarchical context as well and to implement our own agglomerative clustering to test on the OR library instances and help make smarter local merges in the nesting framework.

%\section{Acknowledgements}
%We would like to acknowledge David P. Williamson and Karthik Sridharan for their valuable feedback on this project.
